High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-05-02 notebooks/eccco.qmd
setting up experiments for real-world data

```{julia}
# Counteractual data:
counterfactual_data = load_circles(2000)
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
```

```{julia}
𝒟x = Normal()
𝒟y = Categorical(ones(2) ./ 2)
sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=batch_size)
clf = JointEnergyClassifier(
    sampler;
    builder=MLJFlux.MLP(hidden=(32, 32, 32,), σ=Flux.relu),
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=[1.0,1.0,1e-1],
        verbosity=10,
    ),
)
```

@@

```{julia}
M = ECCCo.ConformalModel(mach.model, mach.fitresult)
factual_label = 0
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 1
factual = predict_label(M, counterfactual_data, x_factual)[1]

>>>>

```{julia}
# Counteractual data:
counterfactual_data = load_circles(1000)
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
```

```{julia}
𝒟x = Normal()
𝒟y = Categorical(ones(2) ./ 2)
sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=batch_size)
clf = JointEnergyClassifier(
    sampler;
    builder=MLJFlux.MLP(hidden=(32, 32, 32,), σ=Flux.swish),
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=[1.0,1.0,1e-2],
        verbosity=10,
    ),
)
```

@@

```{julia}
M = ECCCo.ConformalModel(mach.model, mach.fitresult)
factual_label =  levels(labels)[1]
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target =  levels(labels)[2]
factual = predict_label(M, counterfactual_data, x_factual)[1]

### 2023-05-02 notebooks/mnist.qmd
setting up experiments for real-world data

```{julia}
function pre_process(x; noise::Float32=0.03f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

```{julia}
# Hyper:
_retrain = false
_regen = false

# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

@@

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)
```

@@

```{julia}
# Plot generated samples:
n_regen = 150
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

>>>>

```{julia}
function pre_process(x; noise::Float32=1/256f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

```{julia}
# Hyper:
_retrain = true
_regen = true

# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,5e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=50,
)

@@

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)

Serialization.serialize(joinpath(output_path,"mnist_architectures.jls"), models)
```

@@

```{julia}
# Plot generated samples:
n_regen = 120
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

### 2023-05-02 notebooks/real_world.qmd
setting up experiments for real-world data

# MNIST

## Anecdotal Evidence

### Examples in Introduction

#### Wachter and JSMA

```{julia}
Random.seed!(1234)

# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x_factual = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.9

# Training params:
T = 100
opt = Flux.Optimise.Adam(0.01)
```

```{julia}
# Search:
generic_generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "you_may_not_like_it.png"))
```

#### REVISE

```{julia}
using CounterfactualExplanations.Models: load_mnist_vae
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
```

```{julia}
# Define generator:
revise_generator = REVISEGenerator(
  opt = opt,
  λ=0.0
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, revise_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data_weak = deepcopy(counterfactual_data)
counterfactual_data_weak.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data_weak, M, revise_generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "surrogate_gone_wrong.png"))
```

### ECCCo

```{julia}
function pre_process(x; noise::Float32=1/256f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

```{julia}
# Hyper:
_retrain = true
_regen = true

# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,5e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=50,
)

@@

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)

Serialization.serialize(joinpath(output_path,"mnist_architectures.jls"), models)
```

@@

```{julia}
# Plot generated samples:
n_regen = 120
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

>>>>

# Real-World Data

```{julia}
# Hyper:
_retrain = false
_regen = false

# Data:
n_obs = 10000
datasets = load_tabular_data(n_obs; drop=:credit_default)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 64
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

@@

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)
```

@@

```{julia}
# Plot generated samples:
n_regen = 150
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)
